Supporting Information

### 2021-09-02 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

---
title: ''
output: word_document
bibliography: Zotero_references.bib #This can be generated by exporting BibTex file from Zotero
csl: new-phytologist.csl 
nocite: |
  @mau_temperate_2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006; @chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @roberts_stomatal_1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slot_photosynthetic_2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies. In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. 
Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

@@

Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2). However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.  In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019a], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].

@@

**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Table 1, Table 2).

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018]. In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].

@@

**Leaf photosynthetic thermal tolerance**
Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020]. Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2020]. Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotPhotosyntheticPlasticityTropical2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotPhotosyntheticPlasticityTropical2021]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019a]. The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

Warming impacts forests, from leaf metabolism to whole forest structure. Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$. Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018a; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021]. A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling.  Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019]. Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [McDowell et al. 2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].

>>>>

---
title: ''
output: word_document
bibliography: Zotero_references.bib #This can be generated by exporting BibTex file from Zotero
csl: new-phytologist.csl 
nocite: |
  @mauTemperateTropicalForest2018; @cobleLightDrivesVertical2014; @sackHowStrongIntracanopy2006;@chinLeafAcclimationLight2017; @wykaResponsesLeafStructure2012; @athertonSpatialVariationLeaf2017; @kenzoHeightrelatedChangesLeaf2015; @kusiPlasticLeafMorphology2020; @dangProfilesPhotosyntheticallyActive1997;@gebauerEffectsProlongedDrought2015; @marencoLeafTraitPlasticity2017; @kafutiFoliarWoodTraits2020; @vanwittenbergheVariabilityStomatalConductance2012; @zhang_photosynthetic_2012; @weerasingheCanopyPositionAffects2014; @oldhamHydrostaticGradientNot2010; @ichieEcologicalDistributionLeaf2016; @gregoriouEffectsReducedIrradiance2007; @levizouNondestructiveAssessmentLeaf2005; @liakouraTrichomeDensityIts1997; @fauset_differences_2018; @niinemetsAcclimationHighIrradiance1998; @millenLeafAngleAdaptive1979; @smithShootStructuralEffects1988; @hadleyInfluenceKrummholzMat1987; @panditharathnaChangeLeafStructure2008; @baltzerLeafOpticalResponses2005; @dietzVerticalPatternsDuration2007; @scartazzaInvestigatingEuropeanBeech2016; @duursmaVerticalCanopyGradients2006; @harleyEffectsLightTemperature1996; @hernandezSimilarTemperatureDependence2020; @turnbullScalingFoliarRespiration2003; @chenLeafEconomicsSpectrum2020; @vandewegPhotosyntheticParametersDark2012; @cavaleri_foliar_2008; @konigerXanthophyllCyclePigmentsPhotosynthetic1995; @matsubaraSunshadePatternsLeaf2009; @harrisChangesLeafProperties2013; @hansenVariationPigmentComposition2002a; @poorterLeafOpticalProperties1995; @cobleHowVerticalPatterns2016; @niinemetsCanopyGradientsLeaf2004; @poorterLeafOpticalProperties2000; @taylorNewFieldInstrument2021; @harleyEnvironmentalControlsIsoprene1997; @niinemetsHowLightTemperature2015; @sharkeyFutureIsopreneEmission2014; @simpragaVerticalCanopyGradient2013; @zwienieckiHydraulicLimitationsImposed2004; @sackLeafVenationStructure2013; @ballMaintenanceLeafTemperature1988; @robertsStomatalBoundaryLayerConductances1990; @ambroseEffectsHeightTreetop2010; @zweifelMiddayStomatalClosure2002; @slotPhotosyntheticHeatTolerance2019; @urbanInductionPhotosynthesisImportance2007; @carterWithinCanopyExperimentalLeaf2018a; @martin_boundary_1999; @kosugiSeasonalFluctuationsTemperature2006; @niinemetsVariabilityLeafMorphology2015; @bachofenLightVPDGradients2020; @hamerlynckPhotosyntheticStomatalResponses1996a; @rijkersEffectTreeHeight2000a; @millerOnlySunlitLeaves2021; @legnerWithincanopyVariationPhotosynthetic2014; @kitaoCanopyNitrogenDistribution2018; @rey-sanchez_spatial_2016; @mullerEvidenceEfficientNonevaporative2021; @curtis_intracanopy_2019; @meirLeafRespirationTwo2001; @arakiVerticalSeasonalVariations2017; @bolstad_foliar_1999; @xuSeasonalVariationTemperature2006; @sackHydrologyLeavesCoordination2003
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

**Following the principles outlined above, vertical microclimate profiles, leaf traits, and $g_s$ together explain *in situ* $T_{leaf}$ observations.** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$,* and therefore differs little between understory and canopy.
In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopy than buffered lower-canopies. In some cases canopy leaves exceed the optima for photosynthesis ($T_{opt}$) and $T_{air}$ because of $g_s$-limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower LAI, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light transmission, vertical air mixing, and still air in the lower canopy and heating from the ground [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. 
Greater $RH$ in the inner canopy can also increase $T_{leaf}$ relative to upper canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

@@

Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table 2). However, because of the high VPD and irradiance experienced by sun leaves,  midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests.  In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. The temperature at which $g_s$ is maximized, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021].

@@

**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table 2; @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].**
This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Table 1, Table 2).

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018]. In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021].

@@

**Leaf photosynthetic thermal tolerance**
Irreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death, occurs at $T_{leaf}$ of $40-60^{\circ}C$ [@bakerChlorophyllFluorescenceProbe2008; @feeley_thermal_2020]. Thermal tolerance is described in terms of the temperature at which efficiency of photosystem II starts to decrease, $T_{crit}$, and the $T_{leaf}$ at which efficiency of photosystem II had decreased by 50%, $T_{50}$ [@slotLeafHeatTolerance2021]. Thermal tolerance varies globally across latitude and climate, where $T_{50}$ and $T_{crit}$ decrease across latitudes, and with elevation among tropical forests [@feeley_thermal_2020;  @slotLeafHeatTolerance2021; @osullivanThermalLimitsLeaf2017].

Leaf thermal tolerances are more closely adapted to microclimate than macroclimate [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. Chinquapin oak (*Quercus muehlenbergii*) growing in xeric, sunnier conditions showed higher $T_{50}$ by $2{^\circ}C$ than bur oak (*Quercus macrocarpa*) growing in shaded mesic conditions [@hamerlynckPhotosyntheticStomatalResponses1996a]. In Australia, along a vertical gradient within the canopy, *Acacia papyrocarpa* (Benth.) showed greater $T_{50}$ and higher composite climate stress in the lower, north-facing canopy than other canopy positions, correlating with low wind-speed, greater radiation and $T_{air}$, and lower rates of heat dissapation, as indicated by longer leaf thermal time constants [@curtis_intracanopy_2019]. Additionally, thermal tolerances are explained more by leaf traits than phylogeny [@feeley_thermal_2020; @slotLeafHeatTolerance2021]. $T_{crit}$ was found to be greater among tropical species with large leaves that had slow heating and cooling time, whereas $T_{50}$ was greater among species with greater LMA [@slotLeafHeatTolerance2021]. Across species sun leaves that experienced higher maximum temperatures show higher photosynthetic heat tolerance [@perezPhotosyntheticHeatTolerances2020]. However, considering sun and shade leaf differences within canopy (Table 1), $T_{50}$ varies modestly, being slightly lower for shade than sun leaves for two of three tree species in Panama [@slotPhotosyntheticHeatTolerance2019]. The small difference in $T_{50}$ coupled with large differences in $T_{leaf}$ across thermal microenvironments implies that more thermally tolerant leaves tend to operate closer to their thermal limits and could therefore be more vulnerable to heat anomalies [@perezPhotosyntheticHeatTolerances2020].

@@

Warming impacts forests, from leaf metabolism to whole forest structure. Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$. Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021]. A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling.  Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019]. Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [McDowell et al. 2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].
